The Hypothesis of Testing: Paradoxes arising out of reported coronavirus case-counts

### 2020-05-11 write-up/covid-writeup.tex
covid paper updated

\usepackage[foot]{amsaddr}
\usepackage{biblatex}
%\bibliographystyle{apalike}

@@

\begin{abstract}
Many statisticians, epidemiologists, economists and data scientists have registered serious reservations regarding the reported coronavirus case-counts. Limited testing capacity across many states has been widely identified as a key driver of suppressed coronavirus case-counts.  Calls for increased testing capacity are well-justified and increasingly frequent.  While expanded testing is a laudable goal, selection bias will impact estimates of disease prevalence and the effective reproduction number until the entire population is sampled.  Moreover, tests are imperfect.  False positive/negative rates interact in complex ways with selection bias.  In this paper, we attempt to clarify this interaction.  Through simple calculations, we demonstrate pitfalls and paradoxes that can arise when considering case-count data in the presence of selection bias and measurement-error. The discussion guides a series of suggestions to improve current case-count reporting.
\end{abstract}

@@

\section{Introduction}
The World Health Organization has declared the coronavirus disease 2019 (COVID-19) a public health emergency.  As of April 27th, 2020, a total of 2,993,00 cases have been confirmed worldwide.  As of that afternoon, the New York Times reports at least 965,214 people across the United States have tested positive for the virus, and at least 49,465 patients with the virus have died.  Aggressive policies had been put in place across the US with at least 50\% of the US population officially urged to stay home via state-wide executive actions.

Despite these steps, the data landscape for understanding COVID-19 remains limited.  Public databases maintained by \href{https://bit.ly/2UqFSuA}{Johns Hopkins University} (cite) and the \href{https://bit.ly/2vUHfrK}{New York Times} (cite) provide incoming county-level information of confirmed cases and deaths.  Statisticians, epidemiologists, economists, and data scientists have all used this granular data to forecast COVID-19 case-counts, deaths, and hospitalizations (cite).  Variations on the SIR models (cite) have been used to draw inferences about infection rates and intensity.  Parameter inference most often uses observed case-count and/or observed COVID-19 related deaths to infer latent trajectories of the pandemic (cites).

The goal of this paper is to express reservations at the use of case-counts as a proxy for disease prevalence and intensity as well as its use as direct input into estimation of standard epidemiological models for inference and forecasting.  The reason is simple: current models do not account simultaneously for selection bias and measurement-error. Selection bias enters due to differences in testing strategies across countries and states.  In the US, for example, limited testing capacity has caused local and state health departments to focus on testing only high-risk populations.  Moreover, testing requires the individual to self-select into receiving the test.  Demands for increased testing capacity, while laudable, ignore the issues of self-selection and measurement-error. While increasing testing capacity increases data quantity, there is no guarantee for increased data quality.  We will show that aggressive pushes for ramped up testing capacity that are tied to decreases in data quality may have a very mitigated impact  in our ability to estimate quantities of interest such as prevalence and effective reproduction number.

In this paper, we will demonstrate the complex interaction between these two fundamental concepts.  Measurement-error is often associated with parameter attenuation; we show, however, that the combination with selection effects can cause bias in parameter estimates to change sign and increase/decrease in magnitude.  Moreover, the bias may depend on the sampling fraction, prevalence, and population size.  Without a complete understanding of these interations, we are destined to misinterpret case-counts and come to erroneous conclusions.   We will demonstrate why there is a strong need for more precision and care in presenting COVID-19 analyses to the broader research and non-academic communities.

\subsection{Related work}

There has been an abundance..

However, most complain about testing \emph{capacity} (cite Nate Silver), there is clearly issues of data quality.

Measurement-error discussions.
Selection bias . Proponents of this critique suggest well-designed studies.

Stanford study was Ionides.  They however, forgot about Critique \#2: measurement

@@

This article discusses the relationship between three statistical concepts: selection bias, measurement-error, and the long-forgotten population size. We clarify mathematically why the situation is much more complex than it first appears.  Through simple mathematical arguments, we clarify five important issues related to COVID-19 case-count analysis.  First, we show that unadjusted prevalence rates are (unsurprisingly) biased when tests are imperfect. What is surprising is that the direction and magnitude of the bias can vary substantially and interacts with the sampling fraction.  Next we show that data quality for adjusted prevalence rates includes an additional term in the error decomposition that captures the interplay among measurement error, selection bias, and prevalence.  Third, we show that daily trends in rate of positive tests is equally problematic even if testing rates and selection protocols are held constant.  We show that the rate of change in COVID-19 observed case-counts overestimates the true rate of change prior to the peak time and underestimates it immediately after.  This implies a data scientist analyzing observed rates will be overly pessimistic in the early stages of the pandemic, and overly optimistic subsequent to the peak times.  We show that estimates of the effective reproduction number may be similarly impacted.  Fourth, we show that cross-country comparisons are difficult at best with population size, sampling fraction, and data quality all interacting to impact null hypothesis testing.  Fifth, we discuss a sensitivity analysis method for estimating data quality and its use in understanding the potential biases in COVID-19 case-count analyses.  We then end the paper with a discussion of the benefits of randomized testing, where auxiliary information is most relevant, and the distinction between data quality and decision making in the current pandemic.

\section{Analysis of case-count data}

We start with some simple notation.  Let $N$ denote the population size.  For state-level analysis $N$ is the state's total population, while for country-level analysis $N$ is the country's total population.  At a fixed moment in time, let $Y_j$ denote COVID-19 status for the $j$th individual in the population, $j=1,\ldots, N$. Here, like in survey methodology (cite), we treat COVID-19 status as a fixed but unknown quantity of interest. For simplicity, we start by ignoring the dynamic nature of the viral outbreak as well as the fact that individuals can recover from the disease and assume either individual $j$ is COVID-19 positive and $Y_j=1$ or is COVID-19 negative and $Y_j=0$. We also let $I_j \in \{0,1\}$ be an indicator that the individual was selected for testing ($I = 1$) or not ($I=0$).

@@

When the sampling mechanism is non-random, selection effects may cause bias in the above estimates. To better understand this issue, Meng (2019) provided the following intuitive and powerful statistical decomposition of the error between $\bar y$ and the true proportion $\bar Y$
$$
\bar y_n - \bar Y =  \rho_{I, Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_Y.
$$
The first term represents \emph{data quality}, the second \emph{data quantity}, and the third \emph{problem difficulty}. The term $\rho_{I,Y}$ is the empirical correlation between the population values~$\{ Y_j \}_{j=1}^N$ and the selection values $\{ I_j \}_{j=1}^N$.  Under simple random sampling, $E_{\I} [ \rho_{I,Y} ] = 0$, where the expectation is with respect to the selection mechanism~$\I$, so there is no bias.  The SRS variance formula above shows that $E_{\I} [ \rho_{I,Y}^2 ]  = 1/(N-1)$.  The key issue with selective testing is that $E_{\I} [ \rho_{I,Y} ] \neq 0$.  Meng identified this as the fundamental issue that can lead to paradoxes in the analysis of big data.

Here we highlight the core insights relevant to the COVID-19 crisis.  First, comparing the mean-squared error under selection mechanism $\I$ and SRS, we see that
$$
\frac{E_{\I} \left[ (\bar y_n - \bar Y)^2 \right]}{\sqrt{V_{SRS} (\bar Y)}} = (N-1) E_{\I} \left[ \rho_{I,Y}^2 \right].
$$
This points to a troubling and paradoxical situation: the error relative to SRS increases as a function of population size.  Meng terms this the ``Law of Large Populations'' (LLP).  This points to a critical issue in the current media practices in communication of case-count data: two countries with the same testing strategy (i.e., $E_{\I} [\rho_{I,Y} ]$ equal) can yield wildly different estimates due to population size.  Large countries like the US may have similar true prevalence rates as smaller countries like the UK.  Even under similar testing strategies, the mean-squared error in the prevalence rate for the US will have substantially more variation in comparison to SRS.  Comparing the US to the UK, for example, the MSE will increase by a factor of almost five.  Thus conclusions drawn from observed case-count records may be not just wrong, but very wrong.

@@

Second, there have been calls for increased testing.  While important, many conflate increased testing capacity with increased quality of testing.  We will discuss this point further in Section \ref{section:testingcomments}.  As of May 8th, the US has performed $8,412,095$ tests in total.  Figure~\ref{fig:ustests} shows the trajectory of testing per day and positive cases per day in the US.  The US population is roughly $328$ million, meaning the fraction sampled is $f = 0.026$.  After smoothing, the empirical prevalence is $9.1\%$ on May 8th.  Supposing COVID-19 positive individuals are $2$ times more likely to get tested than those individuals who are COVID-19 negative, the question is ``What is the sample size from a SRS that would yield equivalent MSE in the estimated prevalence?'' Using the above MSE result, Meng showed the effective sample size is equal to $n_{eff} \leq \frac{1}{E_I [ \rho_{I,Y}^2 ]} \frac{f}{1-f}$.  Under these parameters, the effective sample size is $14$.  Recent proposals (cite) have suggested increasing testing capacity which may reduce the relative sampling rate.  Supposing it drops to $1.2$ then the effective sample size will increase to $299$.  See Appendix~\ref{section:effss} for the mathematical details and calculations in additional scenarios.   The above calculations point to a sad state of affairs, saying that the effective sample size even in best case scenarios is not better than a small random sample from the population.

@@

First we require some additional notation.  Let $P_j$ be an indicator of measurement error, equal to $1$ when we mismeasure the outcome and $0$ when we observe the true outcome. We suppose this is a stochastic variable that satisfies $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.  If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^\star = Y_j(1-P_j) + (1-Y_j) P_j$.  Suppose disease prevalence was estimated as the fraction who tested positive for COVID-19, i.e., $\bar y_n^\star = \frac{\sum_{i=1}^N I_j Y_j^\star}{\sum_{i=1}^N I_j}$.  We can again investigate the error compared to the true prevalence $\bar Y$ in statistical terms:
$$
\bar y_n^\star - \bar Y = \sqrt{\frac{1-f}{f}} \left[ \rho_{I,Y} \times \sigma_Y + \rho_{I,PZ} \times \sigma_{PZ} + \sqrt{\frac{f}{1-f}}  \left( FP - (FP+FN) \bar Y \right) \right] .
$$
where $Z = 1-2Y$. The first term represents the perfect testing regime, the second term represents the interaction between imperfect testing and selection bias, while the third time represents the bias due to imperfect testing.  From here on, we refer to $\rho_{I,Y}$ as the \emph{true data quality}, and $\rho_{I,PZ}$ as the \emph{observed data quality} that accounts for both selection bias and measurement-error.  In the appendix, we show the sign of $\rho_{I,PZ}$ is the opposite of the sign of $\rho_{I,Y}$, which implies that the observed data quality adds bias in the opposite direction from the true data quality.

We start by considering the first two terms and assess whether the sign of the bias can reverse due to the interaction of measurement-error and selection bias.  To do this, we define the sampling rates differential.  Let $f_1 := \pr (I_J = 1 \mid Y_J = 1)$ and $f_0 := \pr(I_J = 1 \mid Y_J = 1)$ be the sampling rates.  Then $\Delta = f_1 - f_0$ is the sampling rate differential.  Using these terms, we can re-express the first two terms as
$$
\rho_{I,Y} \times \sigma_Y + \rho_{I,PZ} \times \sigma_{PZ} =
\rho_{I,Y} \times \sigma_Y \left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right].
$$
The final term in brackets is the \emph{measurement-error adjustment to data quality} which is a complex function of sampling rate differential, the odds ratio, and the ratio of measurement-error interaction with prevalence and sampling rates interaction with prevalence. Note $\sgn(\Delta) = \sgn(\rho_{I,Y})$ by equation~\ref{eq:binaryrho} so the measurement-error adjustment either shrinks the data quality measure toward zero or reverses its sign.

While prior investigations have noted the interaction between measurement-error and selection bias (cites), the interaction with the sample size relative to the population, i.e., $f$, has largely been ignored.  The above statistical decomposition clarifies the importance of this quantity~$f$.  In particular, note that the statistical error also includes a bias term due to measurement-error and this term increases as the sampled fraction $f$ increases. Therefore, how the first two terms interact with the final term depends on the fraction of the population sampled.  This interaction is complex, but implies that whether the estimate $\bar y^\star_n$ is an overestimate or underestimate is a complicated question due to the relation amongst these three pieces.

@@

The attentive data analyst will recognize the estimator $\bar y_n$ is biased even for simple random samples and, if sensitivity and specificity were known a priori, may suggest the alternative estimator $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$ which is unbiased under simple random sampling. We again wish to express the error $\tilde  y_n - \bar Y$ in statistical terms. In the appendix, we show that the error now can be expressed as
$$
\tilde y_n - \bar Y = \rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
\times \underbrace{\left[ 1 + FP + FN - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right]}_{D_M}.
$$
The first term is the same as before but increased by $(1 + FP + FN)$ to account for the additional uncertainty due to measurement-error.  The second term is the interaction between selection bias and measurement error.  Figure~\ref{fig:heatmap} presents the measurement-error data quality adjustment as a function of the relative frequency and odds ratio.  We see that the adjustment can be both positive and negative as well as greater than or less than zero.

@@

The above argument shows that the effective sample size is affected in a complex manner by the interaction between selection bias and measurement error.  Here, we discuss the impact of the interaction on the effective sample size when testing capacity is tied to measurement-error.  In the current COVID-19 pandemic, there have been well-justified calls for increases in testing capacity.  These increases, however, may come at the cost of increases in false-positive and false-negative rates.  In terms of effective sample size, if the data quality term remains constant, then testing increases can actually hurt effective sample size.  Consider the simplest case where $f$ increases from $0.05$ to $0.1$; suppose this increase was associated with  false positive rate increasing from $0.5$\% to $5$\% and the false negative rate rising from $5\%$ to $20$\%.  Then the standard analysis would suggest a reduction in the MSE by 47\% but the actual decrease is by 26\%.  This implies that sampling $15\%$ of the population rather than $5\%$ will only increase the effective sample size by a factor $2.9$ rather than the expected $4$ fold increase that is expected.  In a worst case scenario, if the false negative rate climbs to $0.30$ and the false positive rate to $0.10$ then the factor effective sample size increases by only a factor of $2.3$.

@@

The data analyst, now least wary of estimating prevalence and total counts, pauses and thinks.  They return shortly thereafter, with a follow-up consideration: \emph{Ok, perhaps estimating prevalence is difficult. Certainly, however, we can estimate the rate of growth.  All I want to know is when we hit the point at which the curve flattens and number of deaths decrease.  That can't be too hard, surely!}

@@

Consider the state of New York.  Figure~\ref{ny-covid-tests} shows the number of tests and positive tests per day in New York state.  Figure~\ref{ny-covid-frac} presents the fraction of tests that were positive per day in New York state.  For cancellation, we would need the expected data qualities to be uncorrelated and equal. Testing appears relatively stable over a week which suggests assuming $\Delta_t$ (i.e., differential in the selection rate) varies slowly over small windows of time is plausible.  Note, however, that constant $\Delta$ does not imply constant data quality (see Appendix \ref{section:effss} for details).  Figure~\ref{ny-covid-tests} shows the rate of testing remains relatively constant in the last few weeks, specifically relative to the total NY population, so $f_t \approx f_{t-1}$ is reasonable. Finally, the measurement-error adjustment is likely to be constant as no new testing procedures have been reported.  This would not be true in states like Michigan, where testing criteria have been expanding every few weeks.  The final component is an adjustment to this differential error based on the statistical error at the first time point.

Figure \ref{fig:ratio-bias} shows a trajectory of the ratio and potential bias under an SIR model for the epidemic dynamis, with state evolutoin given by the following ODE
$$
\begin{aligned}
\frac{\partial S_t}{\partial t} &= - \beta S_t I_t N^{-1}; \quad
\frac{\partial I_t}{\partial t} = \beta S_t I_t N^{-1} - \gamma I_t; \quad
\frac{\partial R_t}{\partial t} = - \gamma I_t \\
\end{aligned}
$$
where $S_t$ is the number of susceptible individuals at time $t$, $I_t$ is the number of infected individuals at time $t$, $N$ is the total poulation size, and $R_t$ is the number of removed (recovered or deceased) individuals at time $t$.   Figure~\ref{fig:fracpop} presents the fraction of population newly infected individuals at each time step for $\beta = 1.4$ and $\gamma = 0.2$.  Figure \ref{fig:ratio-bias} shows the trajectory of $\bar Y_t / \bar Y_{t-1}$.  We see that in each case, the rate of change in prevalence is likely to be  overestimated prior to the peak and then underestimated afterwards.  From a decision making perspective, such biases have potential impact.  First, the over-estimation may give lawmakers and governors more leverage in proposing aggressive actions that reduce prevalence.  Of course, the analysis supports the argument that the estimates based on available data are overly pessimistic. What both sides miss is that the direction of bias is non-constant over time.  After we hit the peak, estimates will likely under-estimate prevalence.  Such bias puts pressure on lawmakers and governors to prematurely relax social distancing measures.  Moreover, the question is ``Can we trust the observed data to let us know if we have reached the peak?''  It appears that the peak time is the easiest to estimate in terms of having minimal error using availability data.  Note, however, that the standard errors on such estimates will be incredibly wide due to the effective sample size small based on our prior discussions.

>>>>

\usepackage[foot]{amsaddr}
% \usepackage{biblatex}
\usepackage{natbib}
%\bibliographystyle{apalike}

@@

\begin{abstract}
Many statisticians, epidemiologists, economists and data scientists have registered serious reservations regarding the reported coronavirus case-counts. Limited testing capacity across many states has been widely identified as a key driver of suppressed coronavirus case-counts.  Calls for increased testing capacity are well-justified and increasingly frequent.  While expanded testing is a laudable goal, selection bias will impact estimates of disease prevalence and the effective reproduction number until the entire population is sampled.  Moreover, tests are imperfect.  False positive/negative rates interact in complex ways with selection bias.  In this paper, we attempt to clarify this interaction.  Through simple calculations, we demonstrate pitfalls and paradoxes that can arise when considering case-count data in the presence of selection bias and measurement-error. The discussion guides several suggestions on how to improve current case-count reporting.
\end{abstract}

@@

\section{Introduction}
The World Health Organization has declared the coronavirus disease 2019 (COVID-19) a public health emergency.  As of April 27th, 2020, a total of 2,993,00 cases have been confirmed worldwide.  As of that afternoon, the New York Times reports at least 965,214 people across the United States have tested positive for the virus, and at least 49,465 patients with the virus have died.  Aggressive policies had been put in place across the US with at least 50\% of the US population officially urged as of late April to stay home via state-wide executive actions.

Despite these steps, the data landscape for understanding COVID-19 remains limited.  Public databases maintained by \href{https://bit.ly/2UqFSuA}{Johns Hopkins University}~\citep{JHU_Lancet} and the \href{https://bit.ly/2vUHfrK}{New York Times}~\cite{NYT} provide incoming county-level information of confirmed cases and deaths.  Statisticians, epidemiologists, economists, and data scientists have used this granular data to forecast COVID-19 case-counts, deaths, and hospitalizations~\citep{Song2020,Ray2020,2020.IHME}.  In many cases, variations on the SIR models have been used to draw inferences about infection rates and intensity.  Parameter inference often uses observed case-count and/or observed COVID-19 related deaths to infer latent trajectories of the pandemic.

The goal of this paper is to express reservations at the use of case-counts as a proxy for disease prevalence and intensity as well as its use as direct input into estimation of standard epidemiological models for inference and forecasting.  The reason is straightforward: current models do not account simultaneously for selection bias and measurement-error. Selection bias enters due to differences in testing strategies across countries and states.  In the US, for example, limited testing capacity has caused local and state health departments to focus on testing only high-risk populations.  Moreover, testing requires the individual to self-select into asking for a test and then receiving approval by their doctor.  Demands for increased testing capacity, while laudable, ignore the issues of self-selection and measurement-error. While increasing testing capacity increases data quantity, there is no guarantee for increased data quality.  We will show that aggressive pushes for ramped up testing capacity that are tied to decreases in data quality may have a mitigated impact  in our ability to estimate quantities of interest such as prevalence and effective reproduction number.

In this paper, we will demonstrate the complex interaction between these two fundamental concepts.  In classical statistics, measurement-error is often associated with parameter attenuation; we will show, however, that the combination with self-selection can cause bias in parameter estimates to change sign and increase/decrease in magnitude.  Moreover, the bias may depend on the sampling fraction, prevalence, and population size.  Without a complete understanding of these interations, we are destined to misinterpret case-counts and come to erroneous conclusions.   We will demonstrate why there is a strong need for more precision and care in presenting COVID-19 analyses to the broader research and non-academic communities.

% \subsection{Related work}

% There has been an abundance..

% However, most complain about testing \emph{capacity} (cite Nate Silver), there is clearly issues of data quality.

% Measurement-error discussions.
% Selection bias . Proponents of this critique suggest well-designed studies.

% Stanford study was Ionides.  They however, forgot about Critique \#2: measurement

@@

This article discusses the relationship between three statistical concepts: selection bias, measurement-error, and the long-forgotten population size. We clarify mathematically why the situation is much more complex than it first appears.  Through simple mathematical arguments, we clarify five important issues related to COVID-19 case-count analysis.  First, we show that unadjusted prevalence rates are (unsurprisingly) biased when tests are imperfect. What is surprising is that the direction and magnitude of the bias can vary substantially and interacts with the sampling fraction.  Next we show that data quality for adjusted prevalence rates includes an additional term in the error decomposition that captures the interplay among measurement error, selection bias, and prevalence.  Third, we show that daily trends in rate of positive tests is equally problematic even if testing rates and selection protocols are held constant.  We show that the rate of change in COVID-19 observed case-counts overestimates the true rate of change prior to the peak time and underestimates it immediately after.  This implies a data scientist analyzing observed rates will be (on average) overly pessimistic in the early stages of the pandemic, and overly optimistic subsequent to the peak times.  We show that estimates of the effective reproduction number may be similarly impacted.  Fourth, we show that cross-country comparisons are difficult at best with population size, sampling fraction, and data quality all interacting to impact null hypothesis testing.  Fifth, we discuss a sensitivity analysis method for estimating data quality and its use in understanding the potential biases in COVID-19 case-count analyses.  We then end the paper with a discussion of the benefits of randomized testing, use of auxiliary information is most relevant, and the distinction between data quality and decision making in the current pandemic.

\section{Analysis of case-count data}
\label{section:casecount}

We start with some simple notation.  Let $N$ denote the population size.  For state-level analysis $N$ is the state's total population, while for country-level analysis $N$ is the country's total population.  At a fixed moment in time, let $Y_j$ denote COVID-19 status for the $j$th individual in the population, $j=1,\ldots, N$. Here, like in survey methodology~\citep{Cochran77}, we treat COVID-19 status as a fixed but unknown quantity of interest. For simplicity, we start by ignoring the dynamic nature of the viral outbreak as well as the fact that individuals can recover from the disease and assume either individual $j$ is COVID-19 positive and $Y_j=1$ or is COVID-19 negative and $Y_j=0$. We also let $I_j \in \{0,1\}$ be an indicator that the individual was selected for testing ($I = 1$) or not ($I=0$).

@@

The above selection mechanisms are random and independent of the outcome of interest. When this is not the case, selection effects may cause bias in the above estimates. To better understand this issue, \cite{Meng2018} provided the following intuitive and powerful statistical decomposition of the error between $\bar y$ and the true proportion $\bar Y$
$$
\bar y_n - \bar Y =  \rho_{I, Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_Y.
$$
The first term represents \emph{data quality}, the second \emph{data quantity}, and the third \emph{problem difficulty}. The term $\rho_{I,Y}$ is the empirical correlation between the population values~$\{ Y_j \}_{j=1}^N$ and the selection values $\{ I_j \}_{j=1}^N$.  Under simple random sampling, $E_{\I} [ \rho_{I,Y} ] = 0$, where the expectation is with respect to the selection mechanism~$\I$, so there is no bias.  The SRS variance formula above shows that $E_{\I} [ \rho_{I,Y}^2 ]  = 1/(N-1)$.  The key issue with selective testing is that $E_{\I} [ \rho_{I,Y} ] \neq 0$.  Meng identified this as the fundamental issue that can lead to paradoxes in the analysis of big data.

Here we highlight the core insights from~\cite{Meng2018} that are relevant to the COVID-19 crisis.  First, comparing the mean-squared error under selection mechanism $\I$ and SRS, we see that
$$
\frac{E_{\I} \left[ (\bar y_n - \bar Y)^2 \right]}{\sqrt{V_{SRS} (\bar Y)}} = (N-1) E_{\I} \left[ \rho_{I,Y}^2 \right].
$$
This points to a troubling and paradoxical situation: the error relative to SRS increases as a function of population size.  \cite{Meng2018} termed this the ``Law of Large Populations'' (LLP).  This points to a critical issue in the current media practices in communication of case-count data: two countries with the same testing strategy (i.e., $E_{\I} [\rho_{I,Y} ]$ equal) can yield wildly different estimates due to population size.  Large countries like the US may have similar true prevalence rates as smaller countries like the UK.  Even under similar testing strategies, the mean-squared error in the prevalence rate for the US will have substantially more variation in comparison to SRS.  Comparing the US to the UK, for example, the MSE will increase by a factor of almost five.  Thus conclusions drawn from observed case-count records may be not just wrong, but very wrong.

@@

Second, there have been calls for increased testing.  While important, many conflate increased testing capacity with increased quality of testing.  We will discuss this point further in Section \ref{section:testingcomments}.  As of May 8th, the US has performed $8,412,095$ tests in total.  Figure~\ref{fig:ustests} shows the trajectory of testing per day and positive cases per day in the US.  The US population is roughly $328$ million, meaning the fraction sampled is $f = 0.026$.  After smoothing, the empirical prevalence is $9.1\%$ on May 8th.  Supposing COVID-19 positive individuals are $2$ times more likely to get tested than those individuals who are COVID-19 negative, the question is ``What is the sample size from a SRS that would yield equivalent MSE in the estimated prevalence?'' Using the above MSE result, Meng showed the effective sample size is equal to $n_{eff} \leq \frac{1}{E_I [ \rho_{I,Y}^2 ]} \frac{f}{1-f}$.  Under these parameters, the effective sample size is $14$.  Recent proposals~\citep{Siddarth2020} have argued in favor of increasing testing capacity which may reduce the relative sampling rate.  Supposing it drops to $1.2$ then the effective sample size will increase to $299$.  See Appendix~\ref{section:effss} for the mathematical details and calculations in additional scenarios.   The above calculations point to a sad state of affairs, saying that the effective sample size even in best case scenarios is not better than a small random sample from the population.

@@

First we require some additional notation.  Let $P_j$ be an indicator of measurement error, equal to $1$ when we mismeasure the outcome and $0$ when we observe the true outcome. We suppose this is a stochastic variable where $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.  If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^\star = Y_j(1-P_j) + (1-Y_j) P_j$.  Suppose disease prevalence was estimated as the fraction who tested positive for COVID-19, i.e., $\bar y_n^\star = \frac{\sum_{i=1}^N I_j Y_j^\star}{\sum_{i=1}^N I_j}$.  We can again investigate the error compared to the true prevalence $\bar Y$ in statistical terms:
$$
\bar y_n^\star - \bar Y = \sqrt{\frac{1-f}{f}} \left[ \rho_{I,Y} \times \sigma_Y + \rho_{I,PZ} \times \sigma_{PZ} + \sqrt{\frac{f}{1-f}}  \left( FP - (FP+FN) \bar Y \right) \right] .
$$
where $Z = 1-2Y$. The first term represents the perfect testing regime, the second term represents the interaction between imperfect testing and selection bias, while the third time represents the bias due to imperfect testing.  From here on, we refer to $\rho_{I,Y}$ as the \emph{true data quality}, and $\rho_{I,PZ}$ as the \emph{observed data quality} that accounts for both selection bias and measurement-error.  In the appendix, we show the sign of $\rho_{I,PZ}$ is the opposite of the sign of $\rho_{I,Y}$, which implies that the observed data quality adds error in the opposite direction from the true data quality.

We start by considering the first two terms and assess whether the sign of the bias can reverse due to the interaction of measurement-error and selection bias.  To do this, we define the sampling rates differential.  Let $f_1 := \pr (I_J = 1 \mid Y_J = 1)$ and $f_0 := \pr(I_J = 1 \mid Y_J = 1)$ be the sampling rates.  Then $\Delta = f_1 - f_0$ is the sampling rate differential.  Using these terms, we can re-express the first two terms as
$$
\rho_{I,Y} \times \sigma_Y + \rho_{I,PZ} \times \sigma_{PZ} =
\rho_{I,Y} \times \sigma_Y \left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right].
$$
The final term in brackets is the \emph{measurement-error adjustment to data quality} which is a complex function of sampling rate differential, the odds ratio, and the ratio of measurement-error interaction with prevalence and sampling rates interaction with prevalence. Note $\sgn(\Delta) = \sgn(\rho_{I,Y})$ by equation~\ref{eq:binaryrho} in the appendix.  This implies the measurement-error adjustment either shrinks the data quality measure toward zero or reverses its sign.

While prior investigations have noted the interaction between measurement-error and selection bias~\citep{Beesley2020,Beesley2019,Smeden2019}, the interaction with the sample size relative to the population, i.e., $f$, has largely been ignored.  The above statistical decomposition clarifies the importance of this quantity~$f$.  In particular, note that the statistical error also includes a bias term due to measurement-error and this term increases as the sampled fraction $f$ increases. Therefore, how the first two terms interact with the final term depends on the fraction of the population sampled.  This interaction is complex, but implies that whether the estimate $\bar y^\star_n$ is an overestimate or underestimate is a complicated question due to the relation amongst these three pieces.

@@

The attentive data analyst will recognize the estimator $\bar y_n$ is biased even for simple random samples and, if sensitivity and specificity were known a priori, may suggest the alternative estimator $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$ which is unbiased under simple random sampling. We again wish to express the error $\tilde  y_n - \bar Y$ in statistical terms. In the appendix, we show that the error now can be expressed as
\begin{equation}
\label{eq:error}
\rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
\times \underbrace{\left[ 1 + FP + FN - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right]}_{D_M}.
\end{equation}
The first term is the same as before but increased by $(1 + FP + FN)$ to account for the additional uncertainty due to measurement-error.  The second term is the interaction between selection bias and measurement error.  Figure~\ref{fig:heatmap} presents the measurement-error data quality adjustment as a function of the relative frequency and odds ratio.  We see that the adjustment can be both positive and negative as well as greater than or less than zero.

@@

The above argument shows that the effective sample size is affected in a complex manner by the interaction between selection bias and measurement error.  Here, we discuss the impact of the interaction on the effective sample size when testing capacity is tied to measurement-error.  In the current COVID-19 pandemic, there have been well-justified calls for increases in testing capacity.  These increases, however, may come at the cost of increases in false-positive and false-negative rates.  In terms of effective sample size, if the data quality term remains constant, then testing increases may not significantly improve the effective sample size.  Consider the simplest case where $f$ increases from $0.05$ to $0.1$; suppose this increase was associated with  false positive rate increasing from $0.5$\% to $5$\% and the false negative rate rising from $5\%$ to $20$\%.  Then the standard analysis would suggest a reduction in the MSE by 47\% but the actual decrease is by 26\%.  This implies that doubling the sampling rate will only increase the effective sample size by a factor $2.9$ rather than the expected $4$ fold increase that is expected.  In a worst case scenario, if the false negative rate climbs to $0.30$ and the false positive rate to $0.10$ then the factor effective sample size increases by only a factor of $2.3$.

@@

The data analyst, now wary of estimating prevalence and total counts, pauses and thinks.  They return shortly thereafter, with a follow-up consideration: \emph{Ok, perhaps estimating prevalence is difficult. Certainly, however, we can estimate the rate of growth.  All I want to know is when we hit the point at which the curve flattens and number of deaths decrease.  That can't be too hard, surely!}

@@

Consider the state of New York.  Figure~\ref{fig:ny-covid-test} shows the number of tests and positive tests per day in New York state.  Figure~\ref{fig:ny-covid-frac} presents the fraction of tests that were positive per day in New York state.  For cancellation, we would need the expected data qualities to be uncorrelated and equal. Testing appears relatively stable over a week which suggests assuming $\Delta_t$ (i.e., differential in the selection rate) varies slowly over time may be a plausible assumption.  Note, however, that constant $\Delta_t$ does not imply constant data quality (see Appendix \ref{section:effss} for details).  Figure~\ref{fig:ny-covid-test} shows the number of tests has remained relatively constant in the last few weeks, specifically relative to the total NY population, assuming $f_t$ also varies slowly in time is reasonable. Finally, the measurement-error adjustment is likely to be constant as no new testing procedures have been reported.  This would not be true in states like Michigan, where testing criteria have been expanding every few weeks.  The final component is an adjustment to this differential error based on the statistical error at the first time point.

Figure \ref{fig:ratio-bias} shows a trajectory of the true ratio and the potential biased estimators under an SIR model for the epidemic dynamics, with state evolution given by the following ODE
$$
\begin{aligned}
\frac{\partial S_t}{\partial t} &= - \beta S_t I_t N^{-1}; \quad
\frac{\partial I_t}{\partial t} = \beta S_t I_t N^{-1} - \gamma I_t; \quad
\frac{\partial R_t}{\partial t} = - \gamma I_t \\
\end{aligned}
$$
where $S_t$ is the number of susceptible individuals at time $t$, $I_t$ is the number of infected individuals at time $t$, $N$ is the total poulation size, and $R_t$ is the number of removed (recovered or deceased) individuals at time $t$.   Figure~\ref{fig:fracpop} presents the fraction of population newly infected individuals at each time step for $\beta = 1.4$ and $\gamma = 0.2$.  In each case, the rate of change in prevalence is likely to be  overestimated prior to the peak and then underestimated afterwards.  From a decision making perspective, such biases have potential impact.  First, the over-estimation may give lawmakers and governors more leverage in proposing aggressive actions that reduce prevalence.  Of course, the analysis supports the argument that the estimates based on available data are overly pessimistic. What both sides miss is that the direction of bias is non-constant over time.  After we hit the peak, estimates will likely under-estimate prevalence.  Such bias puts pressure on lawmakers and governors to prematurely relax social distancing measures.  Moreover, the question is ``Can we trust the observed data to let us know if we have reached the peak?''  It appears that the peak time is the easiest to estimate in terms of having minimal error using availability data.  Note, however, that the standard errors on such estimates will be incredibly wide due to the low effective sample size as discussed in Section~\ref{section:casecount}.
